Cannibalism among the supermassive black holes

### 2018-06-01 paper.tex
Fixed up more of Introduction. Added papers from Ostriker on mergers and galaxy size.

\section{Introduction}\label{sec:introduction}
What is the ultimate fate of the supermassive black holes (SMBH) resident at the centers of all massive galaxies?  For the galaxies themselves we know that mergers are frequent from the observed (ADD VAN DOKKUM PAPER) evolution of the size and mass of these systems.  But assuming that all galaxy mergers lead to SMBH mergers overpredicts the observations of gravitational waves from the pulsar timing arrays (PTAs) (\citet{2015ApJ...799..178K}, PRETONOS & OSTRIKER, SESANA, NATARJAN).  But we do not often see multiple SMBHs at the centers of massive ellipticals.  So, what does happen?  What we argue in this paper is that as a result of the N-body interactions among the infalling black holes some are ejected, some remain in extended orbits and some (the few most massive ones) do in fact merge but we do not exceed the PTA limits, and the orbiting ones should be detectable via their ability to accrete gas and emit radiation.

A majority of all galaxies with masses $>10^8\msun$ at $z=0$ are believed to host super-massive black holes (SMBHs) in their cores. In a $\Lambda$CDM universe, galaxies at the high end of the galaxy-mass function are merger products of numerous such massive galaxies. Even if only a small fraction of merging galaxies hosted an SMBH in their core at the time of merging, it is reasonable to assume that, during a Hubble time of continuous growth, these galaxies will have substantial phases in which they contain two or more SMBHs at the same time.

Here we look at the merger history of XX exemplary galaxies across the galaxy mass spectrum extracted from a cosmological simulation of hierarchical structure formation. We investigate how, after merging with incoming galaxies, SMBHs diffuse into the cores of the hosts and interact with the resident black hole. We show that gravitational interactions of multiple SMBHs are most probable in high-mass galaxies with $10^{12} \msun < M < 10^{13} \msun$. Galaxies with lower masses  have too few mergers with SMBH hosting galaxies. Galaxies with higher masses are more extended, making dynamical friction processes less efficient and hence failing to drive SMBHs into the host galaxy core.  Several papers (\citet{2018arXiv180403143I}, \citet{2014ApJ...789..156M}) have shown that if all galaxy mergers lead to black hole mergers, the resulting PTA signal would be roughly twice the current limits.  One paper (\citet{2018MNRAS.473.3410R}) has indicated that dynamical interactions among multiple black holes, which eject a non-negligible fraction of the mass, may solve this problem.  The present paper also addresses this purported dynamical solution.

This paper is organized as follows: in Section \ref{sec:methods} we are going to describe the cosmological simulation from which we use the merger history to set up our idealized numerical simulations. We present the few-body integration code, \textsc{AR-Chain} that we used for our simulations of SMBH dynamics, and the modifications we made to this code in order to deal with a host galaxy's gravitational potential. In Section \ref{sec:results}, we show the results of our four exemplary simulations of galaxies growing with time and acquiring new SMBHs. We analyze how the SMBHs are driven into the core of their new host galaxies and how interaction with the host black hole leads to near-ejections or mergers. The final Section \ref{sec:conclusions} contains a discussion of the result and our conclusions.

>>>>

\section{Introduction}\label{sec:introduction}
A majority of all galaxies with masses $>10^8\msun$ at $z=0$ are believed to host super-massive black holes (SMBHs) in their cores.  What is the ultimate fate of the supermassive black holes (SMBH) resident at the centers of these massive galaxies?  For the galaxies themselves we know that mergers are frequent from the observed (VAN DOKKUM PAPERS??) evolution of the size and mass of these systems.  But assuming that all galaxy mergers lead to SMBH mergers overpredicts the observations of gravitational waves from the pulsar timing arrays (PTAs) (\citet{2015ApJ...799..178K}, \citet{2014ApJ...789..156M}, \citet{2008MNRAS.390..192S}, \citet{2013MNRAS.433L...1S}, \citet{2018ApJ...856...42S}, \citet{2009MNRAS.394.2255S}, \citet{2018arXiv180403143I}, NATARAJAN??).  But we do not often see multiple SMBHs at the centers of massive ellipticals.  So, what does happen?  One paper (\citet{2018MNRAS.473.3410R}) has indicated that dynamical interactions among multiple black holes, which eject a non-negligible fraction of the mass, may solve this problem.  The present paper also addresses this purported dynamical solution.  What we argue in this paper is that as a result of the N-body interactions among the infalling black holes some are ejected, some remain in extended orbits and some (the few most massive ones) do in fact merge but we do not exceed the PTA limits, and the orbiting ones should be detectable via their ability to accrete gas and emit radiation.

Here we look at the merger history of XX exemplary galaxies across the galaxy mass spectrum extracted from a cosmological simulation of hierarchical structure formation. We investigate how, after merging with incoming galaxies, SMBHs diffuse into the cores of the hosts and interact with the resident black hole. We show that gravitational interactions of multiple SMBHs are most probable in high-mass galaxies with $10^{12} \msun < M < 10^{13} \msun$. Galaxies with lower masses have too few mergers with SMBH hosting galaxies. Galaxies with higher masses are more extended, making dynamical friction processes less efficient and hence failing to drive SMBHs into the host galaxy core.

This paper is organized as follows: in Section \ref{sec:methods} we describe the cosmological simulation from which we use the merger history to set up our idealized numerical simulations. We present the few-body integration code, \textsc{AR-Chain} that we used for our simulations of SMBH dynamics, and the modifications we made to this code in order to deal with a host galaxy's gravitational potential. In Section \ref{sec:results}, we show the results of our XX exemplary simulations of galaxies growing with time and acquiring new SMBHs. We analyze how the SMBHs are driven into the core of their new host galaxies and how interaction with the host black hole leads to near-ejections or mergers. The final Section \ref{sec:conclusions} contains a discussion of the result and our conclusions.
